---
title: "Report: Attribute analysis"
output: 
  pdf_document:
    latex_engine: pdflatex
  html_document:
    toc: true
    toc_float: true
    code_folding: show
always_allow_html: yes
df_print: kable
params:
  outfolder: ""
  pdata: ""
  covclass: ""
  refcov: ""
---
In this document are reported the result of the analysis of the samples attributes reported in the **`r basename(params$pdata)`** file.

```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
knitr::opts_chunk$set(echo = TRUE)
library("dplyr")
library("ggplot2")
library("reshape")
library("ggpubr")
library("viridis")
```

```{r reading, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# Output folder creation

dir.create(paste0(params$outfolder, "/02_Sample_Data_Analysis"))
of <- paste0(params$outfolder, "/02_Sample_Data_Analysis/")

# Table containing the sample attributes
phenoData <- read.delim(params$pdata, check.names = F)

# List reporting the class of each sample attribute
covariates <- read.delim(params$covclass)

# Identification of the levels of the reference sample attribute
refcov <- params$refcov
cov_class <- levels(phenoData[,refcov])

# Definition of all possible combinations of the reference sample attribute levels
comparisons <- combn(cov_class, m=2)
totest_p <- list()
for(i in 1:ncol(comparisons)){totest_p[[i]] <- comparisons[,i] }
```

```{r setclass, echo=FALSE, message=FALSE, warning=FALSE, results='asis'} 
#### Defining the class of the sample attribute
  for(i in 1:ncol(phenoData)){
    if(as.character(covariates[,2][i]) == "factor"){
      phenoData[, i] = as.factor(phenoData[, i])}
    else{
      class(phenoData[, i]) = as.character(covariates[,2][i])}
  }
```

```{r compade, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# Check reference attribute
if(refcov %in% names(phenoData)){
  cat(paste0("The ", refcov, " attribute is included in the sample attributes.\n\n"), sep="")}else{
  cat(paste0("The ", refcov, " attribute is not included in the sample attributes.\n\n"))
  stop("The indicated reference attribute was not found")}
```

## Analysis among all the sample classes

**`r ncol(phenoData)`** attributes are analysed considering **`r length(cov_class)`** classes of the **`r refcov`** attribute which is used as reference. The attributes are initially analysed using chi-square test for categorical attibutes while ANOVA or Kruskal-Wallis test is applied on continuous attributes. The selection of the test applied on continuous attributes is based on the result of the normality analysis performed using the Shapiro-Wilk test.

```{r testdiff, echo=FALSE, message=FALSE, warning=FALSE, results='asis'} 

# Definition of the output data
out <- matrix(NA, nrow=ncol(phenoData), ncol = length(cov_class)+1)
colnames(out) <- c(cov_class, "p.value")
rownames(out) <- names(phenoData)

# Analysis cycled on each sample attribute
for(j in 1:ncol(phenoData)){
    
    totest <- data.frame(Data = phenoData[,j], refcov = phenoData[,refcov], check.names=F)
    totest <- totest[complete.cases(totest),]  
  
    # Chi-square analysis and summary of categorical attributes
    if(class(phenoData[,j]) == "factor"){
      
      # Chi-square test
      out[j,-1] <- chisq.test(table(totest$Data,totest$refcov))$p.value
      
      # Summary of the attribute classes distribution
      for(i in 1:length(cov_class)){
        class <- cov_class[i]
        tosum <- subset(totest, totest$refcov == class)
        out[j,i] <- toString(paste(names(summary(tosum$Data)), summary(tosum$Data), sep="="))
        }}
    
    # Kruskal-Wallis or ANOVA test and summary for numerical attributes    
    if(class(phenoData[,j]) == "numeric" | class(phenoData[,j]) == "double" | class(phenoData[,j]) == "integer"){
      
      # Normality test
      normaltest <- shapiro.test(totest$Data)$p.value
      
      # Kruskal-Wallis test for non-normally distributed attributes
      if(normaltest < 0.05){out[j,-1] <- kruskal.test(totest$Data~totest$refcov)$p.value}
      
      # ANOVA for normally distributed attributes
      else{out[j,-1] <- anova(lm(totest$Data~totest$refcov))$'Pr(>F)'[1]}
      
      # Summary of the numerical attributes
      for(i in 1:length(cov_class)){
        class <- cov_class[i]
        tosum <- subset(totest, totest$refcov == class)
        avg1 <- round(mean(tosum$Data),2)
        sd1 <- round(sd(tosum$Data),2)
        out[j,i] <- paste(avg1, sd1, sep=" +/- ")}}}

res <- data.frame(Attribute=row.names(out), out, check.names=F, row.names = NULL)
res <- res %>% filter(Attribute != refcov)

res$p.value <- as.numeric(levels(res$p.value))[res$p.value]

# Output writing
write.table(res, paste0(of,"Attribute_analysis_Table_Differential_attribute_among_groups.tsv"), sep="\t", quote=F, row.names=FALSE)
significant <- res %>% filter(p.value < 0.05)

if(nrow(significant) > 0){ 
  out <- paste0("The following **", nrow(significant), "** attribute(s) is/are significantly different (p-value < 0.05) among the sample classes:")}else{
  out <- paste0("None sample attribute is significantly different (p-value < 0.05) among the sample classes.")}
```

```{r covplot, echo=FALSE, message=FALSE, warning=FALSE} 

# Definition of a function for export the plot
printout <- function(name, plot){
    pdf(file=name)
    print(plot)  
    dev.off()
}

# Definition of a function for the stacked bar plot
stackedbar <- function(dat, covid, inlinep=FALSE){
  p <- ggplot(data=dat, aes(x=Var1, y=Freq, fill=Var2)) + geom_bar(stat="identity") + labs(y="Fraction of sample", x="Sample classes", fill=covid) + theme_bw() + scale_fill_viridis(discrete=T)
  if(inlinep==TRUE){print(p)}
  printout(paste0(of, "Attribute_analysis_Plot_Stacked_barplot_", covid, ".pdf"), p)}

# Definition of a function for the grouped bar plot
groupedbar <- function(dat, covid, inlinep=FALSE){
    p <- ggplot(data=dat, aes(x=Var1, y=Freq, fill=Var2)) + geom_bar(stat="identity", position = position_dodge(0.9)) + labs(y=covid, x="Sample classes", fill=covid) + theme_bw() + scale_fill_viridis(discrete=T)
  if(inlinep==TRUE){print(p)}
  printout(paste0(of, "Attribute_analysis_Plot_Grouped_barplot_", covid, ".pdf"), p)}

# Definition of a function for the box plot
boxp <- function(dat, covid, refcov, inlinep=FALSE){
  
  newlimit <- max(dat$value, na.rm = T) + IQR(dat$value, na.rm = T)
  
  p <- ggplot(dat, aes(x=dat[,1], y=value, fill=dat[,1]))+
  geom_violin(trim=FALSE, alpha=0.5) + 
  geom_jitter(position = position_jitter(0.2), aes(fill=dat[,1]), shape=21) + 
  geom_boxplot(width = 0.05, fill="white", outlier.alpha = 0) +
  labs(y=paste0("log10(", covid, ")"), x=refcov, fill=refcov)  + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5)) +
  expand_limits(y = newlimit) + 
  scale_fill_viridis(discrete=T) + 
  stat_compare_means(comparisons = totest_p, label = "p.signif", method = "wilcox.test", hide.ns=F)
  
  if(inlinep==TRUE){print(p)}
  printout(paste0(of, "Attribute_analysis_Plot_Boxplot_", covid, ".pdf"), p)}

#Definition of a function for plotting a set of attributes
checkandplot <- function(dat, n=1, inline=FALSE){
  
  for(j in 1:n){
  covid <- as.character(dat$Attribute[j])
  totest <- phenoData[,c(refcov, covid)]
  
  # Bar plot of categorical attributes
  if(class(totest[,2]) == "factor"){
    toplot <- data.frame(table(totest[,1],totest[,2]))    
    toplot_rel <- data.frame(table(totest[,1],totest[,2])/rowSums(table(totest[,1],totest[,2])))
    stackedbar(toplot_rel, covid, inlinep=inline)
    groupedbar(toplot_rel, covid, inlinep=inline)}
  
  # Violin and Box plot of continuous attributes
  if(class(totest[,2]) == "numeric" || class(totest[,2]) == "integer" || class(totest[,2]) == "double"){
    
    totest[,2] <- log10(totest[,2]+1)
    mm <- melt(totest, id.vars = refcov, na.rm=T)
    boxp(mm, covid, refcov, inlinep=inline)}}
}

if(nrow(significant) > 0){
# In line plot of the three most significant attributes
significant <- arrange(significant, p.value)

# Output plots for all the significant attributes
checkandplot(significant, n=nrow(significant), inline=FALSE)
}
```
`r out`

```{r printtop1, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
if(nrow(significant) > 0){
  cat("The following plot reports the result for the most significantly different attribute.  \n  \n")
  checkandplot(significant, n=1, inline=TRUE)}
```
\pagebreak

## Analysis between pair of sample classes

**`r ncol(phenoData)`** attributes are analysed considering **`r ncol(comparisons)`** pair(s) of sample classes. The attributes are analysed using chi-square test for categorical attibutes while t-test or Wilcoxon Rank-Sum test is applied on continuous attributes. The selection of the test applied on continuous attributes is based on the result of the normality analysis performed using the Shapiro-Wilk test.

```{r pairedanalysis, echo=FALSE, message=FALSE, warning=FALSE, results='asis'} 
# The analysis is cycled on each pair of the reference attribute levels

for(i in 1:ncol(comparisons)){
  
  cov1 = comparisons[1,i]
  cov2 = comparisons[2,i]
  
  # Data subsetting
  phenoData_part <- subset(phenoData, phenoData[refcov] == cov1 | phenoData[refcov] == cov2)
  phenoData_part[, refcov] <- factor(phenoData_part[, refcov])
  
  # Definition empty output data
  pvalue <- rep(NA, ncol(phenoData))
  Summary_cov1 <- rep(NA, ncol(phenoData))
  Summary_cov2 <- rep(NA, ncol(phenoData))
  
  # Definition the colname of the output data
  names(pvalue) <- colnames(phenoData)
  names(Summary_cov1) <- colnames(phenoData)
  names(Summary_cov2) <- colnames(phenoData)
  
  # Analysis cycled on each covariate
  for(j in 1:ncol(phenoData)){
    
    # Selection of the atrribute to test
    totest <- data.frame(Data = phenoData_part[,j], refcov = phenoData_part[,refcov])
    totest <- totest[complete.cases(totest),]  
    totest_c1 <- subset(totest, totest[,"refcov"]==cov1)
    totest_c2 <- subset(totest, totest[,"refcov"]==cov2)
    
    # Chi-square analysis and summary of categorical attributes
    if(class(phenoData[,j]) == "factor"){
      
      # Chi-square test
      pvalue[j] <- chisq.test(table(totest[,1],totest[,2]))$p.value
      
      # Summary of the attribute classes distribution
      Summary_cov1[j] = toString(paste(names(summary(totest_c1[,1])), summary(totest_c1[,1]), sep="="))
      Summary_cov2[j] = toString(paste(names(summary(totest_c2[,1])), summary(totest_c2[,1]), sep="="))}
    
    # Wilcoxon Rank Sum analysis or t-test and summary for numerical attributes
    if(class(phenoData[,j]) == "numeric" | class(phenoData[,j]) == "double" | class(phenoData[,j]) == "integer"){
      
      # Normality test
      normaltest <- shapiro.test(phenoData[,j])$p.value
      
      # Wilcoxon Rank-Sum test for non-normally distributed attributes
      if(normaltest < 0.05){pvalue[j] <- wilcox.test(totest[,1]~totest[,2])$p.value}
      
      # t-test for normally distributed attributes
      else{pvalue[j] <- t.test(totest[,1]~totest[,2])$p.value}
      
      # Summary of the numerical attributes
      avg1 <- round(mean(totest_c1[,1]),2)
      avg2 <- round(mean(totest_c2[,1]),2)
      sd1 <- round(sd(totest_c1[,1]),2)
      sd2 <- round(sd(totest_c2[,1]),2)
      Summary_cov1[j] = paste(avg1, sd1, sep=" +/- ")
      Summary_cov2[j] = paste(avg2, sd2, sep=" +/- ")}}
  
  # Creation of the output data
  res <- data.frame(Attribute=names(phenoData), Summary_cov1, Summary_cov2, pvalue, check.names=F, row.names = NULL)
  res <- res %>% filter(Attribute != refcov)
  names(res) <- c("Attribute", cov1, cov2, "p.value")

  # Output writing
  write.table(res, paste0(of,"Attribute_analysis_Table_Different_attribues_", cov1, "_vs_", cov2, ".tsv"), sep="\t", quote=F, row.names=FALSE)

significant <- res %>% filter(p.value < 0.05)
  
if(nrow(significant) > 0){
  out <- paste0("  \n  \n**", nrow(significant), "** attribute(s) is/are significantly different (p-value < 0.05) between ", cov1, " and ", cov2, " samples.  \n  \n")}else{
  out <- paste0("  \n  \nNone attribute is significantly different (p-value < 0.05) between ", cov1, " and ", cov2, " samples.  \n  \n")}
  cat(out)
  
if(nrow(significant) > 0){
  significant <- arrange(significant, p.value)
  cat("  \n  \nThe following plot reports the result for the most significantly different attribute between ", cov1, " and ", cov2, " samples:", significant$Attribute[1]," \n  \n")
  checkandplot(significant, n=1, inline=TRUE)}}
```

An output file named **Attribute_analysis_Table_Differential_attributes_(class1)_vs_(class2).tsv** is provided were "class1" and "class2" are substituted with the names of classes compared.

## Output files
The following output files were generated:

* Attribute_analysis_Table_Differential_attributes_among_groups.tsv
* Attribute_analysis_Table_Differential_attributes_(class1)_vs_(class2).tsv (class1, class2 = classes compared)
* Attribute_analysis_Plot_Stacked_barplot_(attribute_id).pdf (attribute_id = attribute plotted)
* Attribute_analysis_Plot_Grouped_barplot_(attribute_id).pdf (attribute_id = attribute plotted)
* Attribute_analysis_Plot_Boxplot_(attribute_id).pdf (attribute_id = attribute plotted)

All the output files were generated in the folder **02_Sample_Data_Analysis**