---
title: "Report: Correlation analysis"
output: 
  pdf_document:
    latex_engine: pdflatex
  html_document:
    toc: true
    toc_float: true
    code_folding: show
always_allow_html: yes
params:
  outfolder: ""
  pdata: ""
  covclass: ""
  rthreshold: ""
  pvalthreshold: ""
date: "`r format(Sys.time(), '%d %B, %Y')`"
---
In this document are reported the result of the correlation analysis of the numerical attributes reported in the **`r basename(params$pdata)`** file.

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, results="asis", comment=NA}
knitr::opts_chunk$set(echo = TRUE)
library("ggplot2")
library("corrplot")
library("Hmisc")
library("dplyr")
library("viridis")
```

```{r reading, echo=FALSE, message=FALSE, warning=FALSE, results="asis", comment=NA}
# Output folder creation
setwd(params$outfolder)
dir.create("02_Sample_Data_Analysis")

# Table containing the sample attributes
phenoData <- read.delim(params$pdata, row.names=1, check.names = F)

# List reporting the class of each sample attribute
covariates <- read.delim(params$covclass)
```

```{r setclass, echo=FALSE, message=FALSE, warning=FALSE, results="asis", comment=NA} 
#### Defining the class of the sample attribute
  for(i in 1:ncol(phenoData)){
    if(as.character(covariates[,2][i]) == "factor"){
      phenoData[, i] = as.factor(phenoData[, i])}
    else{
      class(phenoData[, i]) = as.character(covariates[,2][i])}
  }
```

```{r norm, echo=FALSE, message=FALSE, warning=FALSE, results="asis", comment=NA}
# Isolation of the numeric sample attributes
phenoData_num <- phenoData[,unlist(lapply(phenoData, is.numeric))]
phenoData_num <- phenoData_num %>% select_if(~ length(unique(.)) > 1)

# Definition of function for correlation matrix reformat
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut])}

cat("The correlation analysis was perfomed on **", ncol(phenoData_num), "** numerical attributes.\n\n", sep="")

lab <- c("Row attribute", "Column attribute", "Correlation coefficient", "p-value", "BH-adjusted p-value")
```

```{r functions, echo=FALSE, message=FALSE, warning=FALSE, results="asis", comment=NA}
# Definition of a function for export the plot
printout <- function(name, plot){
    pdf(file=name)
    print(plot)  
    invisible(dev.off())}

# Definition of a function for the scatterplot
scatterplot <- function(dat, id1, id2, refcov, inlinep=FALSE){
  p <- ggplot(data=dat, aes(x=dat[,id1], y=dat[,id2], col=phenoData[,refcov])) + 
    geom_point() +
    labs(y=id2, x=id1, col=refcov) + 
    geom_smooth(method=lm, se=F) +
    theme_bw()
  
  if(inlinep==TRUE){print(p)}
  
  printout(paste0("Correlation_Plot_Scatterplot_", id1,"_", id2,".pdf"), p)}
```

```{r pearson, echo=FALSE, message=FALSE, warning=FALSE, results="asis", comment=NA}
cat("## Pearson correlation analysis\n\n")

# Pearson correlation
res_p <- rcorr(as.matrix(phenoData_num), type="pearson")
out_p <- flattenCorrMatrix(res_p$r, res_p$P)
adj_p <- p.adjust(out_p$p, method="BH")
out_p <- cbind(out_p, adj_p)
names(out_p) <- lab
sig_p <- subset(out_p, out_p$`BH-adjusted p-value` < params$pvalthreshold & abs(out_p$`Correlation coefficient`) > params$rthreshold)
sig_p <- arrange(sig_p, sig_p$`BH-adjusted p-value`, desc(abs(sig_p$`Correlation coefficient`)))

cat("Using the Pearson's correlation method, **", nrow(sig_p), "** pair(s) of attributes were significantly correlated.\n\n", sep="")

if(nrow(sig_p)>0){
cat("The following list reports the significant pairs of attributes associated with a significant result from the correlation analysis using the Pearson's method. The significant results are associated with a correlation coefficent greater than **", params$rthreshold, "** and an adjusted p-value lower than **", params$pvalthreshold, "**.\n\n", sep="")

cat(paste(sig_p[,1], sig_p[,2], sep="\tAND\t"), sep="\n\n")
  
cat("\n\nThe result of this analysis were provided with the files:\n\n **Correlation_Table_Attributes_Pearson_correlation_all.txt**\n\n **Correlation_Table_Attributes_Pearson_correlation_significant.txt**.\n\n", sep="")  
}

write.table(out_p, "Correlation_Table_Attributes_Pearson_correlation_all.txt", quote=F, row.names=F, sep="\t")

write.table(sig_p, "Correlation_Table_Attributes_Pearson_correlation_significant.txt", quote=F, row.names=F, sep="\t")

cat("In the following are reported the scatter plot of the most significant correlation computed using the Pearson's method and as heat map all the correlation coefficients computed. The scatter plots of all the significant correlations are provided as pdf files with name **Correlation_Plot_Scatterplot_cov1_cov2.pdf** where cov1 and cov2 are the name of the correlated attributes. The heat map was provided as pdf file named **Correlation_Plot_Corrplot_Pearson_correlation.pdf**.\n\n", sep="")

scatterplot(phenoData_num, as.character(sig_p[1,1]), as.character(sig_p[1,2]), refcov, inlinep=T)
 
for(i in 1:nrow(sig_p)){
scatterplot(phenoData_num, as.character(sig_p[i,1]), as.character(sig_p[i,2]), refcov, inlinep=F)}
```

```{r pears, echo=FALSE, message=FALSE, warning=FALSE, results="hide", comment=NA}
corrplot(res_p$r, order="hclust", hclust.method = "ward.D2", type="upper", col = viridis::viridis(100), tl.srt=45, tl.col="black", tl.cex = 0.5)
dev.off()

pdf("Correlation_Plot_Corrplot_Pearson_correlation.pdf")
corrplot(res_p$r, order="hclust", hclust.method = "ward.D2", type="upper", col = viridis::viridis(100), tl.srt=45, tl.col="black", tl.cex = 0.5)
dev.off()
```

\pagebreak

```{r spearman, echo=FALSE, message=FALSE, warning=FALSE, results="asis", comment=NA}
cat("## Spearman correlation analysis\n\n")

# Spearman correlation
res_s <- rcorr(as.matrix(phenoData_num), type="spearman")
out_s <- flattenCorrMatrix(res_p$r, res_p$P)
adj_s <- p.adjust(out_s$p, method="BH")
out_s <- cbind(out_s, adj_s)
names(out_s) <- lab
sig_s <- subset(out_s, out_s$`BH-adjusted p-value` < params$pvalthreshold & abs(out_s$`Correlation coefficient`) > params$rthreshold)
sig_s <- arrange(sig_s, sig_p$`BH-adjusted p-value`, desc(abs( sig_s$`Correlation coefficient`)))

cat("Using the Spearman's correlation method, **", nrow(sig_s), "** pair(s) of attributes were significantly correlated.\n\n", sep="")

if(nrow(sig_s)>0){
cat("The following list reports the significant results from the correlation analysis using the Spearman's method. The significant results are associated with a correlation coefficent greater than **", params$rthreshold, "** and an adjusted p-value lower than **", params$pvalthreshold, "**.\n\n", sep="")
    
cat(paste(sig_s[,1], sig_s[,2], sep="\tAND\t"), sep="\n\n")

cat("\n\nThe result of this analysis were provided with the files:\n\n **Correlation_Table_Attributes_Spearman_correlation_all.txt**\n\n **Correlation_Table_Attributes_Spearman_correlation_significant.txt**.\n\n", sep="")  
} 

# Write of the output file
write.table(out_s, "Correlation_Table_Attributes_Spearman_correlation_all.txt", quote=F, row.names=F, sep="\t")

write.table(sig_s, "Correlation_Table_Attributes_Spearman_correlation_significant.txt", quote=F, row.names=F, sep="\t")

cat("In the following are reported the scatter plot of the most significant correlation computed using the Spearman's method and as heat map all the correlation coefficients computed. The scatter plots of all the significant correlations are provided as pdf files with name **Correlation_Plot_Scatterplot_cov1_cov2.pdf** where cov1 and cov2 are the name of the correlated attributes. The heat map was provided as pdf file named **Correlation_Plot_Corrplot_Spearman_correlation.pdf**.\n\n", sep="")

scatterplot(phenoData_num, as.character(sig_s[1,1]), as.character(sig_s[1,2]), refcov, inlinep=T)
 
for(i in 1:nrow(sig_p)){
scatterplot(phenoData_num, as.character(sig_s[i,1]), as.character(sig_s[i,2]), refcov, inlinep=F)}
```

```{r spear, echo=FALSE, message=FALSE, warning=FALSE, results="hide", comment=NA}
corrplot(res_s$r, order="hclust", hclust.method = "ward.D2", type="upper", col = viridis::viridis(100), tl.srt=45, tl.col="black", tl.cex = 0.5)
dev.off()

pdf("Correlation_Plot_Corrplot_Spearman_correlation.pdf")
corrplot(res_s$r, order="hclust", hclust.method = "ward.D2", type="upper", col = viridis::viridis(100), tl.srt=45, tl.col="black", tl.cex = 0.5)
dev.off()
```

## Output files
The following output files were generated:

* Correlation_Table_Attributes_Pearson_correlation_all.txt
* Correlation_Table_Attributes_Pearson_correlation_significant.txt
* Correlation_Table_Attributes_Spearman_correlation_all.txt
* Correlation_Table_Attributes_Spearman_correlation_significant.txt
* Correlation_Plot_Corrplot_Pearson_correlation.pdf
* Correlation_Plot_Corrplot_Spearman_correlation.pdf

All the output files were generated in the folder **02_Attribute_Analysis**
