---
title: "Report: Principal Component Analysis (PCA)"
output: 
  pdf_document:
    latex_engine: pdflatex
  html_document:
    toc: true
    toc_float: true
    code_folding: show
always_allow_html: yes
params:
  outfolder: ""
  pdata: ""
  cdata: ""
  covclass: ""
  refcov: ""
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

In this document are reported the result of the Principal Component Analysis (PCA) performed on the **`r basename(params$cdata)`** file.

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("RColorBrewer")
library("gplots")
library("ggplot2")
library("ggfortify")
library("factoextra")
```

```{r reading, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
# Output folder creation
setwd(params$outfolder)
dir.create("02_Sample_Data_Analysis")

# Table containing the covariates
phenoData <- read.delim(params$pdata, row.names=1, check.names = F)

# Table containing the count data
countData <- read.delim(params$cdata, row.names=1, check.names = F)

# List reporting the class of each sample attribute
covariates <- read.delim(params$covclass)

# Identification of the levels of the reference covariate
cov_class <- levels(phenoData[,params$refcov])
```

```{r setclass, echo=FALSE, message=FALSE, warning=FALSE, results="asis"} 
#### Defining the class of the sample attribute
  for(i in 1:ncol(phenoData)){
    if(as.character(covariates[,2][i]) == "factor"){
      phenoData[, i] = as.factor(phenoData[, i])}
    else{
      class(phenoData[, i]) = as.character(covariates[,2][i])}
  }
```

## PCA analysis using the count data
The input count table was preprocessed by removing attributes associated with a zero count in each sample. The count are then log transformed and PCs were computed using zero-centered and Z-score-transformed data.

```{r pcac, echo=FALSE, message=FALSE, warning=FALSE, results="asis", comment=NA} 
### Preprocessing of the count table
countData <- countData[, row.names(phenoData)]
countData <- round(countData, 0)
countData <- countData[rowSums(countData) > 0, ]

### PC computation
cdatat <- data.frame(t(countData), check.names = F)
cdata.pca<-prcomp(log(cdatat+1,2), center=T, scale=T)

imp <- summary(cdata.pca)$importance

cat("The first principal component explain **", paste0(imp[2,1]*100, "%"),"** of the variance in the data while the second principal component explain **", paste0(imp[2,2]*100, "%"),"** of the variance.\n\n", sep="")

write.table(imp, "PCA_Table_PC_count_data_importance.txt", sep="\t", quote=F, col.names=NA)

cat("The file named **PCA_Table_PC_Count_Data_Importance.txt** was provided. This file report the proportion of variance explained by each principal component.", sep="")

## Generation of the Scree plot
cat("The following Scree plot report the contribution of the first components. This plot was saved as **PCA_Plot_Count_data_screeplot.pdf**. \n \n \n", sep="")

p <- fviz_eig(cdata.pca, ggthem = theme_bw(), main=NULL)
print(p)

pdf("PCA_Plot_Count_data_screeplot.pdf")
print(p)
dev.off()

## Generation of PCA plots colored by the reference attribute
cat("\n\nThe following plot report the result of the PCA colored by the reference sample attribute **", refcov,"**. \n\n", sep="")
    
cov <- phenoData[,refcov]
dat <- data.frame(cbind(cov, cdatat), check.names=F)
names(dat)[1] <- refcov

outplot <- autoplot(cdata.pca, data = dat, colour = refcov, size=3.5) + theme_bw() + theme(legend.position="top")
print(outplot)

## Generation of PCA plots colored by the different attributes
cat("A PCA plot was provided applying a color-code based on each sample attribute. Each file was names **PCA_Plot_PCA_count_data_colored_by_(attribute_id).pdf** where attribute_id indicates the specific sample attribute using for coloring the plot. \n \n", sep="")

for(i in 1:ncol(phenoData)){
	
  cov <- phenoData[,i]
  tempdf <- data.frame(phenoData)
  covid <- names(tempdf)[i]
  covid2 <- names(phenoData)[i]
  dat <- data.frame(cbind(cov, cdatat), check.names=T)
  names(dat)[1] <- covid
    
	pdf(paste0("PCA_Plot_PCA_count_data_colored_by_", covid2, ".pdf"), 7, 7)
	outplot <- autoplot(cdata.pca, data = dat, colour=covid, size=3.5, na.value = "gray") + theme_bw() + theme(legend.position="top") + labs(colour=covid2)
	print(outplot)
	dev.off()}
```

## Output files
The following output files were generated:

* PCA_Table_PC_count_data_importance.txt
* PCA_Plot_Count_data_screeplot.pdf
* PCA_Plot_PCA_count_Data_colored_by_(attribute id).pdf (attribute id = attribute used for the analysis)

All the output files were generated in the folder **02_Attribute_Analysis**