---
title: "Report: Data quality control"
output: 
  pdf_document:
    latex_engine: pdflatex
  html_document:
    toc: true
    toc_float: true
    code_folding: show
always_allow_html: yes
params:
  outfolder: ""
  pdata: ""
  cdata: ""
  refcov: ""
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

In this document are reported the quality control analyses of the sample and the count table.

```{r reading, echo=FALSE, message=FALSE, warning=FALSE}
# Output folder creation
outf = paste0(params$outfolder, "/01_QC_and_Data_Preprocessing/")
dir.create(outf)

# Table containing the covariates
phenoData <- read.delim(params$pdata, row.names=1)

# Table containing the count data
countData <- read.delim(params$cdata, row.names=1)

# Identification of the levels of the reference covariate
refcov <- params$refcov
cov_class <- levels(phenoData[,refcov])
```

```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
# Check if the required packages are installed and load them
knitr::opts_chunk$set(echo = TRUE)

packages <- c("dplyr", "ggplot2")

for(i in 1:length(packages)){
  if(packages[i] %in% rownames(installed.packages()) == FALSE)
    install.packages(packages[i])
}

invisible(lapply(packages, library, character.only = TRUE))
```

## Description of the datasets
The sample table stored in the **`r basename(params$pdata)`** file has **`r ncol(phenoData)`** attributes for **`r nrow(phenoData)`** samples.

The count table stored in the **`r basename(params$cdata)`** file has **`r nrow(countData)`** attributes for **`r ncol(countData)`** samples.

The reference sample attribute is **`r refcov`**. The reference attribute is divided in **`r length(cov_class)`** classes.

```{r checkname, echo=FALSE, message=FALSE, warning=FALSE}
# Control of the sample IDs in the datasets
  checkname=0
  if(ncol(countData) == nrow(phenoData) && sum(row.names(phenoData) %in% names(countData)) == ncol(countData)){
  out = "The sample IDs in the two tables are the same."
  checkname=1}else{
  out = "The sample IDs in the two tables are not the same."
  plack <- row.names(phenoData)[row.names(phenoData) %in% names(countData) == F]
  clack <- names(countData)[names(countData) %in% row.names(phenoData) == F]
  
  if(length(clack) > 0){out=paste(out, "\nThe following ID(s) of the count table is/are not present in the samples table:\n ", paste(clack, collapse = ", "), sep="\n")}
  if(length(plack) > 0){out=paste(out, "\nThe following ID(s) of the sample table is/are not present in the count table\n: ", paste(plack, collapse = ", "), sep="\n")}
  }
```
`r out`

## Check for NA values
The sample and count table were checked for NA values.
```{r nacheck, echo=FALSE, message=FALSE, warning=FALSE}
# Verification of NA content of the sample table
checkna=0
if(sum(is.na(phenoData)) == 0){  
  outp <- "The sample table does not contain NA values."
}else{
  NAcolp <- names(phenoData)[colSums(is.na(phenoData))>0]
  NArowp <- row.names(phenoData[rowSums(is.na(phenoData)) > 0,])
  outp <- paste0("\nThe sample table contains at least one NA value in ", length(NAcolp), " attribute(s) for ", length(NArowp), " sample(s).")
  outp <- paste(outp, "\nThe following attribute(s) in the sample data contain(s) at least one NA value:\n", paste(NAcolp, collapse=", "), sep="\n")
  outp <- paste(outp, "\nThe following sample(s) in the sample data contain(s) at least one NA value:\n", paste(NArowp, collapse=", "), sep="\n") 
}

# Verification of NA content of the count table
if(sum(is.na(countData)) == 0){  
  outc <- "\n\nThe count table does not contain NA values."
  checkna=1
}else{
  NAcolc <- names(countData)[colSums(is.na(countData))>0]
  NArowc <- row.names(countData[rowSums(is.na(countData)) > 0,])
  outc <- paste0("\nThe count table contains NA values in ", length(NArowc), " attribute(s) for ", length(NAcolc), " sample(s).")
  outc <- paste(outc, "\nThe following attribute(s) in the count data contain(s) at least one NA value:\n", paste(NArowc, collapse=", "), sep="\n")
  outc <- paste(outc, "\nThe following sample(s) in the count data contain(s) at least one NA value:\n", paste(NAcolc, collapse=", "), sep="\n") 
}
outhead <- "QC report: NA values analysis\n"
```

`r paste(outp, outc, sep="\n")`

## Check for the presence of non numeric values in the count table
The count table is checked for the presence of non numeric data.
```{r numericheck, echo=FALSE, message=FALSE, warning=FALSE}
checknumeric = 0
if(sum(!sapply(countData, is.numeric)) == 0){  
  outc <- "The count table does not contain non numeric data."
  checknumeric=1
}else{
  
  Numcolc <- names(countData)[sapply(countData,class)=="factor"]
  outc <- paste0("\nThe count table contains non numeric values in ", length(Numcolc), " sample(s).")
  outc <- paste(outc, "\nThe following sample(s) in the count data contain(s) at least one non numeric value:\n", paste(Numcolc, collapse=", "), sep="\n") 
}
```
`r paste(outc)`

## Check for normal distribution
The distribution of the numeric attributes of the sample table were checked for normality using the Shapiro-Wilk test.

```{r norm, echo=FALSE, message=FALSE, warning=FALSE}
# Isolation of the numeric sample attributes
phenoData_num <- phenoData[,unlist(lapply(phenoData, is.numeric))]
phenoData_num <- phenoData_num %>% select_if(~ length(unique(.)) > 1)

# Shapiro-Wilk test to verify the normality distribution of each attribute
resnorm <- unlist(mapply(shapiro.test, phenoData_num)["p.value",])

# Analysis of the normality test results
if(sum(resnorm > 0.05) == 0){
  outnorm <- "\nNone attribute of the sample table is normally distributed."}else{
  Normcolp <- names(phenoData_num)[resnorm > 0.05]  
  outnorm <- paste0("\nThe following attribute(s) of the sample table is(are) normally distributed:\n", paste(Normcolp, collapse=", "), sep="\n")}

outhead <- "Preprocessing report: Normal distribution analysis\n"

# Write of the output file
write.table(paste0(outhead, outnorm), paste0(outf, "Preprocessing_Table_Normal_distribution.txt"), quote=F, row.names=F, col.names=F, sep="\t")
```
`r outnorm`

An output file named **Preprocessing_Table_Normal_distribution.txt** is provided with the summary of this analysis. 

## Check for outliers
The numerical variables of the sample table were checked for possible outliers.
```{r chechoutlier, echo=FALSE, message=FALSE, warning=FALSE}
out <<- "Preprocessing report: Outliers analysis\n\n"
outlier_attr <- c()

# Definition of a function to identify the outliers
outlier_find <- function(dt, var) {
  tot <- sum(!is.na(dt[,var]))
  na1 <- sum(is.na(dt[,var]))
  m1 <- mean(dt[,var], na.rm = T)
  med <- median(dt[,var], na.rm = T)
  outlier <- boxplot.stats(dt[,var])$out
  dt2 <- dt
  dt2[,var] <- ifelse(dt[,var] %in% outlier, NA, dt[,var])
  na2 <- sum(is.na(dt2[,var]))
  m2 <- mean(dt2[,var], na.rm = T)
  
  if(length(outlier > 0)){
  outlier_attr <<- c(outlier_attr, var)
  out <<- paste0(out, "The sample attribute \"", var, "\" presents outliers.",
  "\nOutliers identified: ", na2 - na1, " from ", tot, " observations\n",
  "\nMean with outliers: ", round(m1,3),
  "\nMean without outliers: ", round(m2,3), 
  "\nMedian without outliers: ", round(med,3), "\n\n") } else{
  out <<- paste0(out, "The sample attribute ", var, " does not present an outlier.\n\n")
  }}

# Application of the function on each numerical attributes
for(i in 1:ncol(phenoData_num)){
  outlier_find(phenoData_num, names(phenoData_num)[i])}

if(length(outlier_attr)){
    outoutlier <- paste0("The following sample attribute(s) present at least one outlier:\n", paste(outlier_attr, collapse=", "))
}else{outoutlier <- paste0("None sample attributes present at least one outlier.")}

# Write of the output file
write.table(out, paste0(outf,"Preprocessing_Table_Outlier_analysis.txt"), quote=F, row.names=F, col.names=F, sep="\t")
```
`r outoutlier`

```{r outlierplot, echo=FALSE, message=FALSE, warning=FALSE}
l <- list()
# For attributes with outliers report of their distribution
if(length(outlier_attr)>0){
for(i in 1:length(outlier_attr)){
  var <- outlier_attr[i]
  
  p <- ggplot(phenoData_num, aes(x = phenoData_num[,var])) + geom_density(colour="black", fill="white", alpha=0.8) + geom_rug(alpha=0.05) + theme_bw() + labs(x=var) + ggtitle(paste0("Distribution of the ", var, " attribute"))
  
  pdf(paste0(outf,"Preprocessing_Plot_Outlier_analysis_histogram_", var, ".pdf"))
  print(p)
  dev.off()
}}
```

An output file named **Preprocessing_Table_Outlier_analysis.txt** is provided reporting, if present, the outlier values identified with computation of the mean of each attribute with or without outliers. If present, for each attribute presenting an outlier, an histogram plot names **Preprocessing_Plot_Outlier_analysis_histogram_(attribute).pdf** (attribute = identifier of the attribute considered) was provided.

## Output files
The following output files were generated:

* Preprocessing_Table_Normal_distribution_analysis.txt
* Preprocessing_Table_Outlier_analysis.txt
* Preprocessing_Plot_Outlier_analysis_histogram_(attribute).pdf (attribute = identifier of the attribute considered)

All the output file were generated in the folder **01_QC_and_Data_Preprocessing**